---
title: "Clustering"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r load-packages}
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(stringr)
library(tidymodels) 
library(probably)
library(vip)
library(plotly)
tidymodels_prefer() # Resolves conflicts, prefers tidymodel functions
```

# Data Cleaning
```{r data-cleaning}
executions <- read_csv("executiondata_clean.csv")

executions_clean <- executions %>%
  select(-Name, -County) %>%
  filter(Victim_Count < 50) %>%
  separate(Date, c("Month", "Day", "Year"), sep = "/") %>%
  mutate(VictimMale = case_when(
    str_detect(Victim_Sex, "Male") ~ 1, TRUE ~ 0)) %>% # 1 if there was male victim(s)
  mutate(VictimFemale = case_when(
    str_detect(Victim_Sex, "Female") ~ 1, TRUE ~ 0)) %>% # 1 if there was female victim(s)
  mutate(VictimWhite = case_when(
    str_detect(Victim_Race, 'White') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimLatino = case_when(
    str_detect(Victim_Race, 'Latino') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimBlack = case_when(
    str_detect(Victim_Race, 'Black') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimAsian = case_when(
    str_detect(Victim_Race, 'Asian') ~ 1, TRUE ~ 0)) %>%
  mutate(Electrocution = case_when(
    str_detect(Method, 'Electrocution') ~ 'yes', TRUE ~ 'no')) # make a variable that shows if the person was electrocuted
 

executions_clean <- executions_clean %>% select(-Victim_Sex, -Day, -Victim_Race)
executions_clean$Month <-
  as.numeric(as.character(executions_clean$Month)) # turn month into numeric variable
executions_clean$Year <-
  as.numeric(as.character(executions_clean$Year)) # turn year into numeric variable

  
head(executions)
head(executions_clean)


```

# K-Means Clustering

## Variable Exploration

```{r}
# bar chart of region versus method
ggplot(executions_clean, aes(x = Region, fill = Method)) +
    geom_bar(position = 'fill')
  
# box plot of year versus method
ggplot(executions_clean, aes(x = Method, y = Year)) + 
  geom_boxplot() + 
  theme_classic()

# box plot of age versus method
ggplot(executions_clean, aes(x = Method, y = Age)) + 
  geom_boxplot() + 
  theme_classic()
```

Age does not seem to correlate much with method, and does not improve sum of squared error in kclusts from 1-15 when I tried to use it ~ Note from A'di. 


```{r}
# Select the variables to be used in clustering
executions_sub <- executions_clean %>%
    select(Victim_Count, Year)

# Look at summary statistics of the 2 variables
summary(executions_sub)
```
## Picking *K*

```{r}
set.seed(253)

executions_cluster_ss <- function(k){
    # Perform clustering
    kclust <- kmeans(scale(executions_sub), centers = k)

    # Return the total within-cluster sum of squares
    return(kclust$tot.withinss)
}

# create table and graph number of clusters versus sum of squares 
# to pick k using elbow method
tibble(
    k = 1:15,
    tot_wc_ss = purrr::map_dbl(1:15, executions_cluster_ss)
) %>% 
    ggplot(aes(x = k, y = tot_wc_ss)) +
    geom_point() + 
    labs(x = "Number of clusters",y = 'Total within-cluster sum of squares') + 
    theme_classic()
```

According to the elbow method we would pick k = 4 or 5 (I chose 5)

```{r}
set.seed(100)

# Data-specific function to cluster and calculate total within-cluster SS
execution_cluster_silhouette <- function(k){
    # Perform clustering
    kclust <- kmeans(scale(executions_sub), centers = k)

     ss <- cluster::silhouette(kclust$cluster, dist(scale(executions_sub)))
  
    # Return the silhouette measures
    return(mean(ss[, 3]))
}

# Choose value that MAXIMIZES average silhouette
# table with silhouette values
sil <- tibble(
    k = 2:15,
    avg_sil = purrr::map_dbl(2:15, execution_cluster_silhouette)
) 

# graph cluster versus silhouette
sil %>% 
    ggplot(aes(x = k, y = avg_sil)) +
    geom_point() + 
    labs(x = "Number of clusters",y = 'Average Silhouette') + 
    theme_classic()

# view silhouette table
sil
```

The value that maximized the silhouette was k = 12

## Graph K Clusters and Evaluate

### Elbow Method, K = 5

```{r kclust5}
set.seed(253)

# make the 5 clusters
kclust5 <- kmeans(scale(executions_sub), centers = 5)

# add to dataset
executions_clean <- executions_clean %>%
    mutate(kclust_5 = factor(kclust5$cluster))

# graph it
executions_clean %>%
  ggplot(aes(x = Year, y = Victim_Count, color = kclust_5)) +
  geom_point()

# print the total within-cluster sum of squares
kclust5$tot.withinss

# view where execution methods were in the clusters
executions_clean %>%
  count(Method,kclust_5)

# view how different variables average across clusters
executions_clean %>%
    group_by(kclust_5) %>%
    summarize(across(c(Year, Victim_Count), mean))
```

### Silhouette Method, K = 12

```{r kclust12}
set.seed(100)

# make the 12 clusters
kclust12 <- kmeans(scale(executions_sub), centers = 12)

# add to dataset
executions_clean <- executions_clean %>%
    mutate(kclust_12 = factor(kclust12$cluster))

# graph it
executions_clean %>%
  ggplot(aes(x = Year, y = Victim_Count, color = kclust_12)) +
  geom_point()

# get total within cluster sum of squares
kclust12$tot.withinss

# view where methods of executions were in clusters
executions_clean %>%
  count(Method,kclust_12)

# view how different variables average across clusters
executions_clean %>%
    group_by(kclust_12) %>%
    summarize(across(c(Year, Victim_Count), mean))
```

### Conclusions



# Hierarchical Clustering

## Initial Visualizations
```{r}
ggplot(executions_clean, aes(x = Region, fill = Race)) + 
  geom_bar(position = 'fill')

ggplot(executions_clean, aes(x = Method, fill = Race)) + 
  geom_bar(position = 'fill')

ggplot(executions_clean, aes(y = Year, x = Method)) + 
  geom_boxplot()

ggplot(executions_clean, aes(x = Method, y = Victim_Count)) + 
  geom_boxplot() + 
  theme_classic()
```

## Create Data Subset
```{r}
set.seed(555)

executions_subset <- executions_clean
  # slice_sample(n=50)

# Select the variables to be used in clustering
executions_subs <- executions_subset %>%
    select(Year, Victim_Count)

# Summary statistics for the variables
summary(executions_subs)

# See what variables exist in our subset
ggplot(executions_subset, aes(x = Year, y = Victim_Count, color = Method)) + 
  geom_point() + 
  theme_classic()
```

## Clusterings

```{r}
# Compute a distance matrix on the scaled data
dist_mat_scaled <- dist(scale(executions_subs))

# The (scaled) distance matrix is the input to hclust()
# The method argument indicates the linkage type
hc_complete <- hclust(dist_mat_scaled, method = "complete")
hc_single <- hclust(dist_mat_scaled, method = "single")
hc_average <- hclust(dist_mat_scaled, method = "average")
hc_centroid <- hclust(dist_mat_scaled, method = "centroid")

# Plot dendrograms
plot(hc_complete)
plot(hc_single)
plot(hc_average)
plot(hc_centroid)
```

Complete or centroid looks the best, i used complete

## Interpretations & Evaluation

```{r}
# cut tree into groups
executions_subset <- executions_subset %>%
    mutate(
        hclust_h3 = factor(cutree(hc_complete, h = 3)), # Cut at height (h) 3
        hclust_h5 = factor(cutree(hc_complete, h = 5)) # Cut at height (h) 5
    )
```



```{r}
# height is 3

# See what variables exist in our subset
ggplot(executions_subset, aes(x = Year, y = Victim_Count, color = hclust_h3)) + 
  geom_point() + 
  theme_classic()

# view where methods of executions were in clusters
executions_subset %>%
  count(Method,hclust_h3)

# view how different variables average across clusters
executions_subset %>%
    group_by(hclust_h3) %>%
    summarize(across(c(Year, Victim_Count), mean))


# height is 5

ggplot(executions_subset, aes(x = Year, y = Victim_Count, color = hclust_h5)) + 
  geom_point() + 
  theme_classic()

# view where methods of executions were in clusters
executions_subset %>%
  count(Method,hclust_h5)

# view how different variables average across clusters
executions_subset %>%
    group_by(hclust_h5) %>%
    summarize(across(c(Year, Victim_Count), mean))
```

I used height because k-means already took care of clusters...
