---
title: "HW4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages
```{r load-packages}
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(stringr)
library(tidymodels) 
tidymodels_prefer() # Resolves conflicts, prefers tidymodel functions
```

### Data Cleaning
```{r}
executions <- read_csv("executiondata_clean.csv")

executions_clean <- executions %>%
  select(-Name) %>%
  separate(Date, c("Month", "Day", "Year"), sep = "/") %>%
  mutate(VictimMale = case_when(
    str_detect(Victim_Sex, "Male") ~ 1, TRUE ~ 0)) %>% # 1 if there was male victim(s)
  mutate(VictimFemale = case_when(
    str_detect(Victim_Sex, "Female") ~ 1, TRUE ~ 0)) %>% # 1 if there was female victim(s)
  mutate(VictimMixed = case_when(
    (VictimMale==1) & (VictimFemale==1) ~ 1, TRUE ~ 0)) %>% # 1 if the victim(s) were male and female
  mutate(VictimWhite = case_when(
    str_detect(Victim_Race, 'White') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimLatino = case_when(
    str_detect(Victim_Race, 'Latino') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimBlack = case_when(
    str_detect(Victim_Race, 'Black') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimAsian = case_when(
    str_detect(Victim_Race, 'Asian') ~ 1, TRUE ~ 0)) %>%
  mutate(VictimMixedRace = case_when(
    str_detect(Victim_Race, ',') ~ 1, TRUE ~ 0)) %>%
  mutate(Electrocution = case_when(
    str_detect(Method, 'Electrocution') ~ 'yes', TRUE ~ 'no'))


executions_clean <- executions_clean %>% select(-Victim_Sex, -Day, -Victim_Race)
executions_clean$Month <-
  as.numeric(as.character(executions_clean$Month)) # turn month into numeric variable
executions_clean$Year <-
  as.numeric(as.character(executions_clean$Year)) # turn year into numeric variable
executions_clean$Electrocution <- factor(executions_clean$Electrocution, ordered = FALSE )
  
head(executions)
head(executions_clean)
```


### Research Question

We are looking to build a classification model that will predict the method of execution from our data set. To do this, we will use a logistic regression model and a random forests model. Our main aim is to determine if there is a specific method of execution that is more common in certain regions. 


### Logistic Regression

# Visualizations of Predictive Ability

Determined the variables with the most predictive ability are year, region, and race. 

```{r}
ggplot(executions_clean, aes(x = Region, fill = Electrocution)) +
    geom_bar()

ggplot(executions_clean, aes(x = Electrocution, y = Year)) + 
  geom_boxplot() + 
  theme_classic()

ggplot(executions_clean, aes(x = Race, fill = Electrocution)) + 
  geom_bar(position = 'fill')

ggplot(executions_clean, aes(x = Region, fill = Race)) + 
  geom_bar(position = 'dodge')
```


# Implementing Logistic Regression

```{r}
executions_clean <- executions_clean %>%
  mutate(Method = relevel(Electrocution, ref='no')) #set reference level

# Logistic Regression Model Spec
logistic_spec <- logistic_reg() %>%
    set_engine('glm') %>%
    set_mode('classification')

# Recipe
logistic_rec <- recipe(Electrocution ~ Year + Region + Race, data = executions_clean)

# Workflow (Recipe + Model)
log_wf <- workflow() %>% 
    add_recipe(logistic_rec) %>%
    add_model(logistic_spec) 

# Fit Model
log_fit <- fit(log_wf, data = executions_clean)
```

# Evaluating the Model

```{r}
set.seed(123)
data_cv10 <- vfold_cv(executions_clean, v = 10)

# CV Fit Model
log_cv_fit <- fit_resamples(
    log_wf, 
    resamples = data_cv10,
    metrics = metric_set(sens, yardstick::spec, accuracy, roc_auc),
    control = control_resamples(save_pred = TRUE, event_level = 'second'))  # you need predictions for ROC calculations

collect_metrics(log_cv_fit)
```


### Random Forest

### Results

### Conclusion

